---
title: "Analyse Tree Density"
format: html
editor: visual
---

## Analyses Tree Density

Intro to the ms.

## Import data

Import data from NTNU University museum: Natron. Marc did a pivot table in excel.

```{r Loadpackages, warning=FALSE}
library(tidyverse)
library(readxl)
library(DT)
```

Imported DensityAnno2023.xlsx, with 3 regions all tree taxa and 7 heightclasses

```{r}
Data <- read_excel("data/DensityAnno2023.xlsx")
```

Subset rows (20) and columns (10)

```{r}
DT::datatable(Data[1:20,1:10])
```

## Clean data

```{r}
library(dplyr)
   
#Data <- rename(Data,heightclass=`height class 50cm (categorical)` )
```

## Explore data

Check if the datafile is ok

```{r}
table(Data$Region)
```

Three regions included: total of 34911 rows

```{r}
table(Data$Year, Data$Plot)
```

Plot NW, NE, SW, SE referes to plot names in Hedmark-Akershus and corresponds to OV (NW), OH (NE), NV (SW), NH (SE) in the two other regions.

## Figures

A scatter plot and loess smoother showing the development of larger birch trees over time.

```{r}

```

```{r}
Data %>%
  #filter(Taxa == "Betula pubescens", HeightClass > 5)%>%
  ggplot(aes(x=Year,y=Quantity))+
  geom_jitter(
             size=3,
             width=0.1,
             alpha=0.1)+
  geom_smooth(method = "loess",
              se=F)+
  facet_grid(cols=vars(Treatment),
             rows=vars(Region),
             #ncol=1,
             scales = "free_y")+
  theme_bw(base_size = 12)
  
```

```{r}
Data %>%
  filter(Taxa == "Betula pubescens")%>%
  ggplot(aes(x=Year,y=Quantity))+
  geom_jitter(
             size=3,
             width=0.1,
             alpha=0.1)+
  geom_smooth(method = "loess",
              se=F)+
  facet_grid(cols=vars(Treatment),
             rows=vars(Region),
             #ncol=1,
             scales = "free_y")+
  theme_bw(base_size = 12)
```

```{r}
Data %>%
  filter(Taxa == "Sorbus aucuparia")%>%
  ggplot(aes(x=Year,y=Quantity))+
  geom_jitter(
             size=3,
             width=0.1,
             alpha=0.1)+
  geom_smooth(method = "loess",
              se=F)+
  facet_grid(cols=vars(Treatment),
             rows=vars(Region),
             #ncol=1,
             scales = "free_y")+
  theme_bw(base_size = 12)
```

```{r}
Data %>%
  filter(Taxa == "Pinus sylvestris")%>%
  ggplot(aes(x=Year,y=Quantity))+
  geom_jitter(
             size=3,
             width=0.1,
             alpha=0.1)+
  geom_smooth(method = "loess",
              se=F)+
  facet_grid(cols=vars(Treatment),
             rows=vars(Region),
             #ncol=1,
             scales = "free_y")+
  theme_bw(base_size = 12)
```

```{r}
Data %>%
  filter(Taxa == "Picea abies")%>%
  ggplot(aes(x=Year,y=Quantity))+
  geom_jitter(
             size=3,
             width=0.1,
             alpha=0.1)+
  geom_smooth(method = "loess",
              se=F)+
  facet_grid(cols=vars(Treatment),
             rows=vars(Region),
             #ncol=1,
             scales = "free_y")+
  theme_bw(base_size = 12)
```

We can look at species differences, and at the same time reduce the amount of points by taking the mean per location and treatement.

```{r, warning=F}
Data %>%
  filter(HeightClass > 5,
         Taxa %in% c("Betula pubescens", "Picea abies", "Pinus sylvestris", "Sorbus aucuparia"))%>%
  group_by(LocalityCode, Year, Taxa, Treatment)%>%
  summarise(Quantity = mean(Quantity)) %>%
  ggplot(aes(x=Year,y=Quantity, colour=Treatment, fill=Treatment))+
  geom_jitter(
             size=3,
             width=0.1,
             alpha=0.4)+
  geom_smooth(method = "loess",
              se=F)+
  facet_wrap(.~Taxa)+
  theme_bw(base_size = 12)
```

In the figure above we see little obvious differences in birch, pine and spruce, but large differences in rowan.

Since there is auto correlation (repeated measures at the same locatins every year), we could chose to visualise this as lines arther than points.

```{r, warning=F}
Data %>%
  filter(HeightClass > 5,
         Taxa %in% c("Betula pubescens", "Picea abies", "Pinus sylvestris", "Sorbus aucuparia"))%>%
  group_by(LocalityCode, Year, Taxa, Treatment)%>%
  summarise(Quantity = mean(Quantity)) %>%
  ggplot(aes(x=Year,y=Quantity, group=LocalityCode))+
  geom_line(alpha=0.8)+
  facet_grid(rows = vars(Taxa),
             cols=vars(Treatment))+
  theme_bw(base_size = 12)
```

We could look at the same thing, but for the smalles height class

```{r, warning=F}
Data %>%
  filter(HeightClass ==1,
         Taxa %in% c("Picea abies", "Sorbus aucuparia"))%>%
  group_by(LocalityCode, Year, Taxa, Treatment)%>%
  summarise(Quantity = mean(Quantity)) %>%
  ggplot(aes(x=Year,y=Quantity, group=LocalityCode))+
  geom_line(alpha=0.8)+
  facet_grid(rows = vars(Taxa),
             cols=vars(Treatment),
             scales="free")+
  theme_bw(base_size = 12)
```

```{r, warning=F}
Data %>%
  filter(Year==2023,
         Taxa %in% c("Picea abies", "Sorbus aucuparia"))%>%
  pivot_longer(cols = Quantity) %>%
  ggplot(aes(x=Treatment,y=HeightClass))+
  geom_violin(alpha=0.8,
              fill="grey30")+
  theme_bw(base_size = 12)+
  facet_wrap(.~Taxa)
```

```{r, warning=F}
Data %>%
  filter(Year %in% c(2009, 2013, 2018, 2023),
         Taxa %in% c("Sorbus aucuparia"))%>%
  pivot_longer(cols = Quantity) %>%
  ggplot(aes(x=Treatment,y=HeightClass, fill=Treatment))+
  geom_violin(alpha=0.8,
              adjust = 3
              #fill="grey30"
              )+
  theme_bw(base_size = 12)+
  facet_wrap(.~Year, ncol=4)+
  guides(fill="none")
```

```{r, warning=F}
Data %>%
  filter(Year %in% c(2009, 2013, 2018, 2023),
         Taxa %in% c("Picea abies"))%>%
  pivot_longer(cols = Quantity) %>%
  ggplot(aes(x=Treatment,y=HeightClass, fill=Treatment))+
  geom_violin(alpha=0.8,
              adjust = 3
              #fill="grey30"
              )+
  theme_bw(base_size = 12)+
  facet_wrap(.~Year, ncol=4)+
  guides(fill="none")
```

```{r, warning=F}
Data %>%
  filter(Year %in% c(2009, 2013, 2018, 2023),
         Taxa %in% c("Pinus sylvestris"))%>%
  pivot_longer(cols = Quantity) %>%
  ggplot(aes(x=Treatment,y=HeightClass, fill=Treatment))+
  geom_violin(alpha=0.8,
              adjust = 3
              #fill="grey30"
              )+
  theme_bw(base_size = 12)+
  facet_wrap(.~Year, ncol=4)+
  guides(fill="none")
```

```{r, warning=F}
Data %>%
  filter(Year %in% c(2009, 2013, 2018, 2023),
         Taxa %in% c("Betula pubescens"))%>%
  pivot_longer(cols = Quantity) %>%
  ggplot(aes(x=Treatment,y=HeightClass, fill=Treatment))+
  geom_violin(alpha=0.8,
              adjust = 3
              #fill="grey30"
              )+
  theme_bw(base_size = 12)+
  facet_wrap(.~Year, ncol=4)+
  guides(fill="none")
```

Gunnars plot 0712

```{r}
Data %>%
  filter(Region=='SUSTHERB Telemark',EngelskNavn=='Pine' )%>%
  ggplot()+
  geom_violin(aes(x=factor(Year),
                 y=Quantity
                 ),
              fill = "yellow")+
  facet_wrap(.~Treatment,
             ncol=1,
             scales = "free")+
  theme_bw(base_size = 12)
  
```

Number of Pine at plot level (circles) in Telemark: low spread, but reduction of plots with high abundance from 2017.No temporal trends in UB

```{r}
Data %>%
  filter(Region=='SUSTHERB Telemark',EngelskNavn=='Spruce' )%>%
  ggplot()+
  geom_violin(aes(x=factor(Year),
                 y=Quantity
                 ),
              fill = "yellow")+
  facet_wrap(.~Treatment,
             ncol=1,
             scales = "free")+
  theme_bw(base_size = 12)
```

Number of Spruce at plot level (circles) in Telemark: low spread, but reduction of plots with high abundance from 2016, but more in B vs UB

```{r}
Data %>%
  filter(Region=='SUSTHERB Telemark',EngelskNavn=='Birch' )%>%
  ggplot()+
  geom_violin(aes(x=factor(Year),
                 y=Quantity
                 ),
              fill = "yellow")+
  facet_wrap(.~Treatment,
             ncol=1,
             scales = "free")+
  theme_bw(base_size = 12)
```

Number of Birch at plot level (circles) in Telemark: low spread, but reduction of plots with high abundance in B from 2013.Less temporal trends in UB

```{r}
Data %>%
  filter(Region=='SUSTHERB Telemark',EngelskNavn=='Rowan' )%>%
  ggplot()+
  geom_violin(aes(x=factor(Year),
                 y=Quantity
                 ),
              fill = "yellow")+
  facet_wrap(.~Treatment,
             ncol=1,
             scales = "free")+
  theme_bw(base_size = 12)
```

Number of Rowan at plot level (circles) in Telemark: low spread, but no reduction of plots in B and UB

```{r}
Data %>%
  filter(Region=='SUSTHERB Trondelag')%>%
  ggplot()+
  geom_violin(aes(x=factor(Year),
                 y=Quantity
                 ),
              fill = "yellow")+
  facet_wrap(.~Treatment,
             ncol=1,
             scales = "free")+
  theme_bw(base_size = 10)
```

Number of trees at plot level (circles) in Trondelag: low spread, but reduction of plots with high abundance in B from 2013 and in UB

```{r}
Data %>%
  filter(Region=='SUSTHERB Trondelag',EngelskNavn=='Pine' )%>%
  ggplot()+
  geom_violin(aes(x=factor(Year),
                 y=Quantity
                 ),
              fill = "yellow")+
  facet_wrap(.~Treatment,
             ncol=1,
             scales = "free")+
  theme_bw(base_size = 10)
```

Number of Pine trees at plot level (circles) in Trondelag: low spread, but reduction of plots with high abundance in B and in UB from 2013

```{r}
Data %>%
  filter(Region=='SUSTHERB Trondelag',EngelskNavn=='Spruce' )%>%
  ggplot()+
  geom_violin(aes(x=factor(Year),
                 y=Quantity
                 ),
              fill = "yellow")+
  facet_wrap(.~Treatment,
             ncol=1,
             scales = "free")+
  theme_bw(base_size = 10)
```

Number of Spruce trees at plot level (circles) in Trondelag: low spread, but reduction of plots with high abundance in UB from 2013.Less temporal trends in B

```{r}
Data %>%
  filter(Region=='SUSTHERB Trondelag',EngelskNavn=='Birch' )%>%
  ggplot()+
  geom_violin(aes(x=factor(Year),
                 y=Quantity
                 ),
              fill = "yellow")+
  facet_wrap(.~Treatment,
             ncol=1,
             scales = "free")+
  theme_bw(base_size = 10)
```

Number of Birch trees at plot level (circles) in Trondelag: low spread, but no reduction of plots with high abundance in B from 2013.Increasing temporal trends in UB

```{r}
Data %>%
  filter(Region=='SUSTHERB Trondelag',EngelskNavn=='Rowan' )%>%
  ggplot()+
  geom_violin(aes(x=factor(Year),
                 y=Quantity
                 ),
              fill = "yellow")+
  facet_wrap(.~Treatment,
             ncol=1,
             scales = "free")+
  theme_bw(base_size = 10)
```

Number of rowan trees at plot level (circles) in Trondelag: low spread, but some reduction of plots with high abundance in B and UB

```{r}
Data %>%
  #filter(Region=='SUSTHERB Telemark',EngelskNavn=='Pine' )%>%
  ggplot()+
  geom_violin(aes(x=factor(Year),
                 y=Quantity
                 ),
              fill = "yellow")+
  facet_wrap(.~Treatment,
             ncol=1,
             scales = "free")+
  theme_bw(base_size = 10)
```

Number of pine trees at plot level (circles) in Telemark: low spread,

```{r}
Data %>%
  filter(EngelskNavn=='Pine' )%>%
  ggplot()+
  geom_violin(aes(x=factor(Year),
                 y=Quantity
                 ),
              fill = "yellow")+
  facet_wrap(.~Treatment,
             ncol=1,
             scales = "free")+
  theme_bw(base_size = 10)
```

Number of trees at plot level (circles) across all regions: low spread, but some reduction of plots with high abundance in B and UB from 2013

```{r}
Data %>%
  filter(EngelskNavn=='Spruce' )%>%
  ggplot()+
  geom_violin(aes(x=factor(Year),
                 y=Quantity
                 ),
              fill = "yellow")+
  facet_wrap(.~Treatment,
             ncol=1,
             scales = "free")+
  theme_bw(base_size = 10)
```

Number of spruce trees at plot level (circles) across all regions: low spread, but some reduction of plots with high abundance in B and UB

```{r}
Data %>%
  filter(EngelskNavn=='Birch' )%>%
  ggplot()+
  geom_violin(aes(x=factor(Year),
                 y=Quantity
                 ),
              fill = "yellow")+
  facet_wrap(.~Treatment,
             ncol=1,
             scales = "free")+
  theme_bw(base_size = 10)
```

Number of birch trees at plot level (circles) across all regions: low spread, no reduction of plots with high abundance in B and UB

```{r}
Data %>%
  filter(EngelskNavn=='Rowan' )%>%
  ggplot()+
  geom_violin(aes(x=factor(Year),
                 y=Quantity
                 ),
              fill = "yellow")+
  facet_wrap(.~Treatment,
             ncol=1,
             scales = "free")+
  theme_bw(base_size = 10)
```

Number of rowan trees at plot level (circles) across all regions: low spread, no reduction of plots with high abundance in B and UB

```{r}
Data %>%
  filter(Region=='SUSTHERB Trondelag',EngelskNavn=='Pine' )%>%
  ggplot()+
  geom_point(size=3,alpha=.5,
aes(x=factor(Year),
                 y=Quantity
                 ),
              fill = "yellow")+
  facet_wrap(.~Treatment,
             ncol=1,
             scales = "free")+
  theme_bw(base_size = 10)
```

```{r}
Data %>%
  filter(Region=='SUSTHERB Trondelag',EngelskNavn=='Birch' )%>%
  ggplot()+
  geom_point(size=3,alpha=.5,
aes(x=factor(Year),
                 y=Quantity
                 ),
              fill = "yellow")+
  facet_wrap(.~Treatment,
             ncol=1,
             scales = "free")+
  theme_bw(base_size = 10)
```

```{r}
Data %>%
  filter(Region=='SUSTHERB Trondelag',EngelskNavn=='Spruce' )%>%
  ggplot()+
  geom_point(size=3,alpha=.5,
aes(x=factor(Year),
                 y=Quantity
                 ),
              fill = "yellow")+
  facet_wrap(.~Treatment,
             ncol=1,
             scales = "free")+
  theme_bw(base_size = 10)
```

```{r}
Data %>%
  filter(Region=='SUSTHERB Trondelag',EngelskNavn=='Rowan' )%>%
  ggplot()+
  geom_point(size=3,alpha=.5,
aes(x=factor(Year),
                 y=Quantity
                 ),
              fill = "yellow")+
  facet_wrap(.~Treatment,
             ncol=1,
             scales = "free")+
  theme_bw(base_size = 10)
```

```{r}
Data %>%
  filter(Region=='SUSTHERB Trondelag',EngelskNavn=='Birch',
        HeightClass=="7" )%>%
  ggplot()+
  geom_violin(aes(x=factor(Year),
                 y=Quantity
                 ),
              fill = "yellow")+
  facet_wrap(.~Treatment,
             ncol=1,
             scales = "free")+
  theme_bw(base_size = 10)
```

Number of birch trees at plot level (circles) in Trøndelag at harvest class 7: increased in UB from 2018 and to lesser extent in B from 2019

```{r}
Data %>%
  filter(Region=='SUSTHERB Trondelag',EngelskNavn=='Spruce',
        HeightClass=="7" )%>%
  ggplot()+
  geom_violin(aes(x=factor(Year),
                 y=Quantity
                 ),
              fill = "yellow")+
  facet_wrap(.~Treatment,
             ncol=1,
             scales = "free")+
  theme_bw(base_size = 10)
```

Number of spruce trees at plot level (circles) in Trøndelag at harvest class 7: increased in B from 2016 and to lesser extent in UB from 2019

```{r}
Data %>%
  filter(Region=='SUSTHERB Trondelag',EngelskNavn=='Rowan',
        HeightClass=="7" )%>%
  ggplot()+
  geom_violin(aes(x=factor(Year),
                 y=Quantity
                 ),
              fill = "yellow")+
  facet_wrap(.~Treatment,
             ncol=1,
             scales = "free")+
  theme_bw(base_size = 10)
```

Number of Rowan trees at plot level (circles) in Trøndelag at harvest class 7: increased in UB from 2015. No in B

```{r}
Data %>%
  filter(EngelskNavn=='Rowan',
        HeightClass=="7" )%>%
  ggplot()+
  geom_violin(aes(x=factor(Year),
                 y=Quantity
                 ),
              fill = "yellow")+
  facet_wrap(.~Treatment,
             ncol=1,
             scales = "free")+
  theme_bw(base_size = 10)
```

```{r}
Data %>%
  filter(Region=='SUSTHERB Trondelag',EngelskNavn=='Rowan',
        HeightClass=="1")%>%
  ggplot()+
  geom_violin(aes(x=factor(Year),
                 y=Quantity
                 ),
              fill = "yellow")+
  facet_wrap(.~Treatment,
             ncol=1,
             scales = "free")+
  theme_bw(base_size = 10)
```

Number of rowan trees at plot level (circles) in Trøndelag at harvest class 1:

```{r}
Data %>%
  filter(Region=='SUSTHERB Trondelag',EngelskNavn=='Rowan',
        HeightClass=="2")%>%
  ggplot()+
  geom_violin(aes(x=factor(Year),
                 y=Quantity
                 ),
              fill = "yellow")+
  facet_wrap(.~Treatment,
             ncol=1,
             scales = "free")+
  theme_bw(base_size = 10)
```

```{r}
Data %>%
  filter(Region=='SUSTHERB Trondelag',EngelskNavn=='Rowan',
        HeightClass=="3")%>%
  ggplot()+
  geom_violin(aes(x=factor(Year),
                 y=Quantity
                 ),
              fill = "yellow")+
  facet_wrap(.~Treatment,
             ncol=1,
             scales = "free")+
  theme_bw(base_size = 10)
```

```{r}
install.packages()

 
```

```{r}
Data %>%
mixed.lmer<-lmer(Quantity~Treatment (1|localityCode/Plot))  
  
```
