---
title: "Browsing pressure"
format: html
editor: visual
---

```{r}

library(readxl)
library(knitr)
library(tidyverse)

dat <- read_excel("C:/rwrkd/Github/SustHerbTreeDensity/data/Browsing2023.xlsx")

## changing covariate names to avoid issues
colnames(dat)[34] <- "unbrowsed_count"
colnames(dat)[35] <- "sum_old_new"


out_tot <- NULL #final dataframe with 1 rows = 1 site/year 
for(site in unique(dat$siteNumber)){ #first loop dealing with the site 
  
  #Making a subset of the full dataframe with all individuals (browsed+unbrowsed) from the site and without P abies sp  
  tmp <- dat[dat$siteNumber %in% site & !dat$scientificName %in% "Picea abies",]
   
  
  for(y in unique(tmp$year)){#second loop dealing with the different years for each site
    #create a dataframe to store results for the specific site/year 
    out <- data.frame(site=site, year=y, UB=NA, SON=NA, SAB=NA, exp=NA, local=NA)
    
    #new subset but with only the specific year 
    tmp2 <- tmp[tmp$year==y,]
    #transform unbrowsed count in numeric data
    tmp2$unbrowsed_count <- as.numeric(tmp2$unbrowsed_count)
    
    #depending on the type of experiment (unbrowsed or browsed) the calcul for SAB 
    # (sum of available browsed and browsed) is different so 2 conditions
    if(unique(tmp2$experiment) %in% "Unbrowsed"){ #first condition for unrowsed
      UB <- sum(na.exclude(tmp2$unbrowsed_count)) #sum of available shoots
      SON <- sum(na.exclude(tmp2$sum_old_new)) #sum of browsed (old and new)
      SAB <- sum(na.exclude(tmp2$unbrowsed_count)) # sum of available browsed and browsed == #sum of available shoots
      exp <- "Unbrowsed" 
      local <- unique(tmp2$locality) #need the name of the locality later to make the correspondance between unbrowsed and browsed for one site
    }
    if(unique(tmp2$experiment) %in% "Browsed"){ #second condition for browsed
      UB <- sum(na.exclude(tmp2$unbrowsed_count))#sum of available shoots
      SON <- sum(na.exclude(tmp2$sum_old_new))#sum of browsed (old and new)
      SAB <- sum(na.exclude(tmp2$unbrowsed_count)) + sum(na.exclude(tmp2$sum_old_new)) # this time sum of available browsed and browsed == #sum of available shoots + #sum of browsed (old and new)
      exp <- "Browsed"
      local <- unique(tmp2$locality)
    }
    
    # here we add results in our dataframe for the site/year
    out$UB <- UB 
    out$SON <- SON
    out$SAB <- SAB
    out$exp <- exp
    out$local <- local
    
    #we bind the result out to our final dataframe out_tot that will enventually contains the results for all site/year
    out_tot <- rbind(out_tot, out)
    
  }
  
}

## Now we want for each site and year the dAUC: result store in out_tot2
## here we simplify our dataframe out_tot by locality and year (one row of out_tot2 ## will be one site/year) and we compute the dAUC 
out_tot2 <- out_tot %>% group_by(local, year) %>% summarise(dAUC = out_tot$SAB[out_tot$local %in% local & out_tot$exp %in% "Unbrowsed" & out_tot$year %in% year] - out_tot$SAB[out_tot$local %in% local & out_tot$exp %in% "Browsed" & out_tot$year %in% year])



```

## Browsing index

The index is based on De Vriendt et al. 2020the annual biomass increments inside the exclosires. The allometric models are published in the supplementary information in the first paper. The first paper only had the Trøndelag sites, and therefore it also was able to combine the annual biomass increment with the canopy cover values to get an even better index. The second paper included the Telemark site, but there has not been calculated any productivity indeces for the Hedmark-Akershus sites.

> **Goal** - To calculate a productivity index for the three regions, using all the years of data.

```{r}
#| echo: false
2 * 2
```

The `echo: false` option disables the printing of code (only output is displayed).
